\chapter{Conclusion and Future Work}

\section{Summary of Contributions}

This thesis delivered \TimeLens{}, an AI-powered bilingual guide for the
\GEM{}, through two engineering contributions and a market study. The first is an
\textbf{on-device artifact detector} whose development isolated label quality---not
architecture or resolution---as the decisive factor, yielding a compact YOLOv8n model
that resolves every previously failing class. The second is a \textbf{grounded
bilingual RAG guide}, selected from a seven-model benchmark and optimized from over
30\,s to about 10\,s end-to-end latency, that keeps answers anchored to curated museum
data in English or Arabic. Both ship in a production Flutter application, and a market
analysis situates the system within the heritage-AR landscape.

\section{Critical Reflection}

The decisive gains in this project came not from model complexity but from
\textbf{clean data and grounding}: hand-verified labels made the on-device detector
trustworthy, and retrieval grounding kept the language model factual. The honest
residual risks are equally clear---validation frame leakage and training-to-deployment
domain shift on the detection side, fine-grained long-tail classes, and evolving
real-world capture conditions.

\section{Future Work}

\paragraph{Dataset.} Expand the dataset significantly by capturing and annotating additional museum artifacts scaling the dataset to eventually cover all artifacts in the museum.

\paragraph{Detection.} Further improve the inference speed of the detection model. While the current system is already fast enough to provide a smooth user experience, further optimization could enable more real-time performance and improve the model's ability to track subtle changes in artifact locations and bounding boxes between frames.

\paragraph{Retrieval-augmented guide.} Expand the closed knowledge base; replace or
augment the keyword query classifier with a learned intent model; add conversational
memory across turns; and build index health-audit and maintenance tooling so the
knowledge base can be updated and validated routinely as the collection grows.

\paragraph{System.} Conduct primary user studies (including System Usability Scale evaluation) in live museum
conditions to quantify learning outcomes and long-horizon experience quality.

