\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This thesis documents the design, implementation, and evaluation of \TimeLens{},
an AI- and AR-powered mobile guide for the \GEM{} (GEM). Pointing a phone at an
exhibit, a visitor sees the artifact recognized in real time and can ask follow-up
questions answered in English or Arabic. The work targets three problems specific
to in-gallery deployment: fine-grained visual similarity among \textbf{51 catalogued
artifacts} (many near-identical Ramesside statues), the gap between curated training
data and handheld phone-camera conditions, and the risk of an AI guide stating
historical ``facts'' it cannot support.

The thesis presents two engineering-research contributions and a market study. The
first is an \textbf{on-device artifact detector} developed through a
data-quality-driven iteration study: from foundation-model auto-annotation
(YOLO-World), through spatial label-cleaning rules, to a fully hand-annotated
dataset. The study isolates \emph{label quality}---rather than model architecture or
input resolution---as the decisive factor: the final \textbf{YOLOv8n} model resolves
every previously failing class while remaining a \textbf{5.97\,MB} TensorFlow Lite
asset that runs in real time on a mid-range phone. The second is a \textbf{bilingual
Retrieval-Augmented Generation (RAG) guide} grounded in a \textbf{108-record ChromaDB}
knowledge base; a systematic benchmark of \textbf{seven candidate language models}
selects \textbf{Gemma 4 E2B (Q4\_K\_M)}, and a ten-step optimization reduces
end-to-end latency from \textbf{over 30\,s to about 10\,s}. Both ship in a production
\textbf{Flutter} application with a bilingual interface, museum location gating, and
text-to-speech.

Beyond standard evaluation metrics, the thesis presents the detector's performance during deployment and field testing,
where no misclassifications were observed across the artifact classes included in the dataset and detection confidence scores
consistently ranged between 75\% and 95\%. The document further provides reproducibility notes, an ethics and bias statement, and a formal generative-AI disclosure.
%Beyond headline metrics, the thesis reports a conservative held-out
%result (mAP@0.5 $=0.751$) alongside the in-distribution figures. The document
%includes reproducibility notes, an ethics and bias statement, and a formal
%generative-AI disclosure.

\vspace{1.2em}
\noindent\textbf{Keywords:} on-device object detection; YOLOv8n; TensorFlow Lite;
retrieval-augmented generation (RAG); Gemma; ChromaDB; bilingual question answering;
augmented reality; Flutter; cultural heritage computing; Grand Egyptian Museum.

