\chapter{AI Ethics and Bias Statement}

\section{Data Fairness}
We assess class balance and confusion concentration to detect under-represented
categories and biased behavior. Per-class precision/recall/mAP (recorded in the
detector's \texttt{per\_class\_metrics.csv}) and the normalized confusion matrix expose
the fine-grained classes most prone to error, rather than hiding them behind an average.

\section{Mitigation}
Bias and failure are mitigated primarily through \emph{label quality}: hand annotation
removed systematic auto-labeling errors, augmentation broadens coverage, and per-class
confidence thresholds (e.g., a stricter floor for the false-positive-prone
\emph{Siptah Stela}) reduce confident mistakes. On the language side, retrieval
grounding plus an explicit Arabic name reference reduce hallucination and mis-transliteration.

\section{Transparency and Explainability}
The work reports per-class metrics and the confusion matrix, and openly discloses the
validation frame-leakage and domain-shift limitations together with conservative
held-out numbers. The RAG guide answers only from retrieved, citable museum context.

\section{Ethical Risk Assessment}
The principal risk is misplaced visitor trust in an uncertain prediction. The system
mitigates this with a calibrated confidence gate that suppresses low-confidence
detections rather than guessing, on-device inference that uploads no per-frame imagery,
and a locally served language model that keeps visitor queries out of third-party
services.
